FLM for growth as function of SPEI

### 2021-01-05 analysis/paper/paper.Rmd
added TRMM to spei analysis
Switched models to use SPEI data calculated from Xavier et al. gridded precip.

Monthly precipitation and potential evapotranspiration records ( <!--# date range -->) were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. We calculated 3-month SPEI using the `SPEI` package in \`r version\$version.string\` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. <!--# something about why SPEI-3 is appropriate. -->  Precipitation and evapotranspiration data were obtained from <!--# decide on data source and insert methods -->. We compared the SPEI values calculated using this data source to <!--# others --> including <!--# either ipmuted dataset if using published or famous published if using imputed --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).

>>>>

Monthly precipitation and potential evapotranspiration records were used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. We calculated 3-month SPEI using the `SPEI` package in \`r version\$version.string\` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. <!--# something about why SPEI-3 is appropriate. --> Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. We compared the SPEI values calculated using this data source to <!--# others --> including <!--# TRMM, near-by weather stations, global SPEI --> and the timing of drought events is consistent although the exact magnitude of drought differs among datasets ( <!--# ref to appendix 1 -->).

### 2021-01-05 analysis/paper/vital rate models.Rmd
Switched models to use SPEI data calculated from Xavier et al. gridded precip.

# library(HeliconiaDemography) #load functions in this compendium
# devtools::load_all()
```

@@

- 
- ~~Remove seedlings from dataset, for now~~
- ~~Separate models for CF and fragment, for now~~
- Add model for probability of reproduction and other vital rates
- Model seedlings separately?
- Add random effects, interactions, other complexities
- Match SPEI data to demographic plots programtically by lat lon data

@@

```{r data, echo=TRUE}
spei <- read_rds(here("analysis", "data", "derived_data", "spei_lags.rds"))
xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
```
# Join data sets

## Gridded data
Get only rows for february (the month of the census), and year for joining

@@

Now, for the demography data, get year and plot coords

```{r}
# demog$plot %>% unique() %>% sort()
# plot_coords$plot %>% unique() %>% sort()
demog <-
  right_join(plot_coords %>% select(plot, lon, lat),
             demog,
             by = "plot") %>% 
  select(-x_09, -y_09) #remove to avoid confusion since i'm not using here.
```

Now join by year and whether plot is in grid cell or not.  Probably need to construct `sf` object for grid cells (knowing that the lat lon are the centers and the resolution is 0.25º) and then use some `sf_` function to check if a point is inside of a grid cell.  See: https://stackoverflow.com/questions/59766153/left-join-based-on-closest-lat-lon-in-r

## Imputed data:

`demog$year` = `spei$census_year`

Getting site to match up will take more work.

```{r}
unique(demog$plot)
unique(demog$bdffp_reserve_no)
unique(spei$site)
```

km41 = 1501
km37 = ??
florestal = 1301, Florestal-CF
cabo frio = 3402, CaboFrio-CF
Porto Alegre = 3114, PortoAlegre-CF
Colosso = 1104, 1202
Dimona = 2206, 2108, 2107, Dimona-CF

colosso_clust = 1301, 3402, 1104, 1202, CaboFrio-CF, Florestal-CF

```{r}
demog <- 
  demog %>% 
  mutate(rain_site = case_when(
    bdffp_reserve_no %in% c("1301", "3402", "1104", "1202")      ~ "colosso_clust",
    plot %in% c("CaboFrio-CF", "Florestal-CF")                   ~ "colosso_clust",
    bdffp_reserve_no %in% c("2206", "2108", "2107")              ~ "dimona",
    plot == "Dimona-CF"                                          ~ "dimona",
    bdffp_reserve_no == "3114"                                   ~ "porto_alegre",
    plot == "PortoAlegre-CF"                                     ~ "porto_alegre",
    bdffp_reserve_no == "1501"                                   ~ "km_clust",
    TRUE ~ NA_character_
  ))
```

Get SPEI history only for census months (assuming all censuses done in February)

```{r}
spei_census <- 
  spei %>% 
  filter(month(yearmonth) == 2) %>% 
  mutate(year = year(yearmonth), .before = site) %>% 
  select(-yearmonth)
```

Join

```{r}
demog_full <-
  left_join(demog, spei_census, by = c("rain_site" = "site", "year")) %>% 
  mutate(log_shts = log(shts), .after = shts)

# demog_full %>% select(spei, spei_history)
```

@@

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
```

@@

To simplify initial modeling attempts, I'll split the data into seedlings vs. post-seedlings and by habitat.  I'll build the Q and L matrixes describing exposure and lag for each subset.

@@

```{r}
g_post_cf <-
  gam(shts_next ~ 
        s(log_shts) + #does this *need* to be log # shoots?
        # s(shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
gam.check(g_post_cf) #looks good
# summary(g_post_cf)
anova(g_post_cf)
```

```{r}
g_post_1ha <-
  gam(shts_next ~ 
        s(log_shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
gam.check(g_post_1ha)
anova(g_post_1ha)
```

@@

```{r}
p_surv_cf_slice <-
  pred_s_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_cf_slice
```

```{r}
p_surv_1ha_slice <-
  pred_s_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_1ha_slice
```

# Results

```{r include=FALSE}
#TODO: switch to predict() so y axis is correct
g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])

g_df <-
  bind_rows(
    "CF" = g_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = g_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )
```

I'm not 100% sure this is correct since it's not done with `predict()`

```{r}
g_df %>% 
  ggplot(aes(x = exp(x), y = exp(fit + coef(g_post_cf)[1]))) +
  geom_jitter(data = demog_full %>% filter(habitat %in% c("1-ha", "CF")),
              aes(x = shts, y = shts_next), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = exp(fit + coef(g_post_cf)[1] - se),
                  ymax = exp(fit + coef(g_post_cf)[1] + se)),
              alpha = 0.2,
              fill = "blue") +
  labs(x = "size (#shoots)", y = "size (t + 1)") +
  facet_wrap(~habitat) +
  theme_bw() +
  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
```
Plants don't grow much.

```{r}
anova(g_post_cf)
anova(g_post_1ha)
```

There's a significant effect of lagged drought on number of shoots in the next year.

```{r}
(p_growth_cf | p_growth_1ha) +
  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
  plot_layout(guides = "collect")
```

It looks like drought in October is bad for growth in fragments, but drought in June and July is much worse for plants in continuous forest.

Reminder that these are still fit with two different models and that `predict()` is using different climate histories (`spei_history`).  Might not be entirely comparable?  Different penalties to the smooths make these hard to compare as well.

@@

```{r}
(p_growth_cf_slice + theme(axis.text.x = element_blank())) /
  p_growth_1ha_slice +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
```

@@

```{r include=FALSE}
s_cf <- plot.gam(s_post_cf, shift = coef(s_post_cf)[1])
s_1ha <- plot.gam(s_post_1ha, shift = coef(s_post_1ha)[1])

s_df <-
  bind_rows(
    "CF" = s_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = s_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )
```

I'm not 100% sure this is correct since it's not done with `predict()`

```{r}
s_df %>% 
  ggplot(aes(x = exp(x), y = plogis(fit + coef(s_post_cf)[1]))) +
  geom_line() +
  geom_ribbon(aes(ymin = plogis(fit + coef(s_post_cf)[1] - se),
                  ymax = plogis(fit + coef(s_post_cf)[1] + se)),
              alpha = 0.2) +
  labs(x = "size (#shoots)", y = "P(survival)") +
  facet_wrap(~habitat) +
  stat_summary_bin(data = demog_post %>% filter(habitat %in% c("1-ha", "CF")),
                   geom = "point",
                   fun = mean,
                   bins = 20,
                   aes(y = surv, x = shts)) +
  theme_bw() +
  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
```

```{r}
anova(s_post_cf)
anova(s_post_1ha)
```

@@

```{r}
(p_surv_cf|p_surv_1ha) + 
  plot_annotation(title = "Effect of drought on survival") +
  plot_layout(guides = "collect")
```

The surface for 1-ha fragments is basically flat except for a slight *negative* effect of recent wet weather on survival.  In continuous forest, drought in the previous February and March has a negative impact on survival.  Wet conditions or very dry conditions in the previous August–October have a positive effect on survival.

@@

```{r}
(p_surv_cf_slice / p_surv_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
```

>>>>

library(HeliconiaDemography) #load functions in this compendium
# devtools::load_all()
```

```{r}
#define plotting themes
theme_talk <- theme_bw() + theme(text = element_text(size = 18))
```

@@

- 
- ~~Remove seedlings from dataset, for now~~
- ~~Separate models for CF and fragment, for now~~
- Add model for probability of reproduction and other vital rates
- Model seedlings separately?
- Add random effects, interactions, other complexities
- Try integrative measure of size.  Some paper you read did some formula with height and leaf number. Plantago?

@@

```{r data, echo=TRUE}
xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
```

```{r}
length(unique(demog$ha_id_number))
range(demog$year)
2009-1998
```

# Join data sets

## Prep precip data
Get only rows for february (the month of the census), and year for joining

@@

Figure out which sites correspond to which grid cells.  The `xa` lat lon are the centers of grid cells that are 0.25º^2

```{r}
unique(xa2$lat)
mean(c(-2.25, -2.5))
```
So, from -2.25 to -2.5

```{r}
unique(xa2$lon)
mean(c(-59.5, -59.75))
mean(c(-59.75, -60)) # Florestal-CF, 5752, 5751, 5750, 5756, CaboFrio-CF
mean(c(-60, -60.25)) # 2206, 2108, 2107, Dimona-CF, 5753, PortoAlegre-CF
```

```{r}
plot_coords
```

```{r}
demog2 <-
  demog %>% 
  mutate(lon = case_when(
    plot %in% c("Florestal-CF", "5752", "5751", "5750", "5756", "CaboFrio-CF") ~ "-59.875",
    plot %in% c("2206", "2108", "2107", "Dimona-CF", "5753", "PortoAlegre-CF") ~ "-60.125",
    TRUE ~ NA_character_
  ), .before = plot)
demog_full <- 
  full_join(demog2, xa2, by = c("lon", "year")) %>% 
  select(-lon, -lat, -yearmonth)
```

## Tidy

```{r}
demog_full <-
  demog_full %>% 
  mutate(log_shts = log(shts), .after = shts) %>% 
  mutate(log_shts_next = log(shts_next), .after = shts_next)
```

@@

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
  mutate(flowered = as.numeric(infl > 0), .after = infl)
# View(demog_full)
```

@@

Because I can't model interactions with DLNMs, I can either include factors like habitat as predictors (different intercepts, same functional form, same data) or fit separate models (potentially different functional forms, but not using all the data).

@@

```{r}
g_post_cf <-
  gam(shts_next ~ 
        s(log_shts) + #does this *need* to be log # shoots?
        # s(year, bs = "re") + #random effect of year
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

Inspect the GAM

```{r}
gam.check(g_post_cf) 
```

Suggests possibly more knots for s(log_shts), but it looks like a straight line to me

```{r}
plot(g_post_cf, pages=1)
```

```{r}
summary(g_post_cf)
```

Remember, R-sq is probably so high because size this year is a very good predictor of size next year.

```{r}
g_post_1ha <-
  gam(shts_next ~ 
        s(log_shts) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "log"),
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
gam.check(g_post_1ha)
```

Looks great

```{r}
summary(g_post_1ha)
```

@@

```{r}
p_surv_cf_slice <-
  pred_s_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_surv_cf_slice
```

```{r}
p_surv_1ha_slice <-
  pred_s_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(survival)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_surv_1ha_slice
```

## Flowering Probability

### Fit GAMs
```{r}
x <- demog_post_cf$flowered
sum(x, na.rm = TRUE) / length(x)
```

```{r}
f_post_cf <-
  gam(flowered ~ 
        s(log_shts) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
# gam.check(f_post_cf)
```

```{r}
summary(f_post_cf)
```

```{r}
f_post_1ha <-
  gam(flowered ~ 
        s(log_shts) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
summary(f_post_1ha)
```

```{r}
pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

```{r}
flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
                  max(pred_f_post_1ha$fitted, pred_f_post_cf$fitted))
plogis(flowered_lim)
```

```{r}
p_flowered_cf <-
  pred_f_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_flowered_cf
```
Something is wrong.  Flowering probability should not be 1.

```{r}
p_flowered_1ha <-
  pred_f_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_flowered_1ha
```

```{r}
p_flowered_cf_slice <-
  pred_f_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(flowering)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_cf_slice
```

```{r}
p_flowered_1ha_slice <-
  pred_f_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(flowering)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_1ha_slice
```

# Plots
## Growth
```{r}
p_growth_spei <-
  (p_growth_cf | p_growth_1ha) +
  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
  plot_layout(guides = "collect") & theme_talk
p_growth_spei
ggsave(here("growth_spei_xa.png"), p_growth_spei, height = 7, width = 14)
```

It looks like drought in September/October is bad for growth.

Reminder that these are still fit with two different models and might not be entirely comparable.  Different penalties to the smooths make these hard to compare as well.

@@

```{r}
(p_growth_cf_slice + theme(axis.text.x = element_blank())) /
  p_growth_1ha_slice +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
ggsave(here("growth slice_xa.png"))
```

@@

@@

```{r}
p_surv_spei <- 
  (p_surv_cf|p_surv_1ha) + 
  plot_annotation(title = "Effect of drought on survival") +
  plot_layout(guides = "collect") & theme_talk
p_surv_spei
ggsave(here("surv_spei_xa.png"), p_surv_spei, width = 14, height = 7)
```

The surface for 1-ha fragments is basically flat except for a slight *negative* effect of recent wet weather on survival.  In continuous forest, wet conditions or very dry conditions in the previous August–October have a positive effect on survival.

@@

```{r}
((p_surv_cf_slice + theme(axis.text.x = element_blank())) / p_surv_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
ggsave(here("survival slice_xa.png"))
```
Pretty flat.

## Flowering

```{r}
p_flowered_spei <- 
  (p_flowered_cf|p_flowered_1ha) + 
  plot_annotation(title = "Effect of drought on flowering") +
  plot_layout(guides = "collect") & theme_talk
p_flowered_spei
ggsave(here("flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
```

The surface for continuous forest is perfectly flat (did I do something wrong with the color scale?) and for 1-ha fragments there is a strange (anomalous?) negative effect of high SPEI (wet conditions) in the summer 2 years ago.

```{r}
((p_flowered_cf_slice + theme(axis.text.x = element_blank())) / p_flowered_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
ggsave(here("flowering slice.png"))
```
